---
title: "Homework 5 - Advanced Data in R"
format: docx
---

```{r}
#| include: false
library(gapminder)
data(gapminder)

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

library(dplyr)
library(tidyr)
```

Fill out [this Quarto document](https://github.com/BIFX552-22/CourseInfo/hw/05advanced_data.qmd) while completing the lesson on writing data, `dplyr` and `tidyr`.

### [Writing Data](https://swcarpentry.github.io/r-novice-gapminder/11-writing-data/index.html)

::: callout-tip
## Key Points

-   Save plots from RStudio using the 'Export' button.

-   Use `write.table` to save tabular data.
:::

#### Challenge 1

Modify the code below to print a second page in the pdf, showing a facet plot (hint: use `facet_grid`) of the same data with one panel per continent.

```{r}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)

# Page 1
ggplot(gapminder, aes(year, lifeExp, color=country)) +
  geom_line() +
  theme(legend.position = "none")

# Page 2
ggplot(gapminder, aes(year, lifeExp, color=country)) +
  geom_line() +
  theme(legend.position = "none") +
  facet_grid(~ continent)

# You then have to make sure to turn off the pdf device!

dev.off()
```

#### Challenge 2

Write a data-cleaning script file that subsets the gapminder data to include only data points collected since 1990.

Use this script to write out the new subset to a file in the `cleaned-data/` directory.

```{r}
sub90 <- gapminder[gapminder$year >= 1990,]
write.csv(sub90, 'recent.csv')
head(sub90)
```

### [Data frames with `dplyr`](https://swcarpentry.github.io/r-novice-gapminder/13-dplyr/index.html)

::: callout-tip
## Key Points

-   Use the `dplyr` package to manipulate data frames.
-   Use `select()` to choose variables from a data frame.
-   Use `filter()` to choose data based on values.
-   Use `group_by()` and `summarize()` to work with subsets of data.
-   Use `mutate()` to create new variables.
:::

#### Challenge 1

Write a single command (which can span multiple lines and includes pipes) that will produce a data frame that has the African values for `lifeExp`, `country` and `year`, but not for other Continents. How many rows does your data frame have and why?

```{r}
subAfrica <- filter(gapminder, continent == 'Africa') %>%
  select(lifeExp, country, year) %>%
  rename(life_expectancy = lifeExp)
  # filter(continent == 'Europe' & year == 2007)
  # Already subsetted data before piping it, no need to filter again.
str(subAfrica)
nrow(subAfrica)
```

#### Challenge 2

Calculate the average life expectancy per country. Which has the longest average life expectancy and which has the shortest average life expectancy?

```{r}
gapsum <- gapminder %>%
  group_by(country) %>%
  summarize(avg_lifeExp = mean(lifeExp))

# Lowest avg lifeExp
filter(gapsum, avg_lifeExp == min(avg_lifeExp))

# Highest avg lifeExp
filter(gapsum, avg_lifeExp == max(avg_lifeExp))

gapsum_map <- gapminder %>%
  group_by(year, continent) %>%
  summarize(avg_lifeExp = mean(lifeExp))

ggplot(gapsum_map, aes(year, avg_lifeExp, color = continent)) + geom_line()
  

```

#### Challenge 3

Calculate the average life expectancy in 2002 of 2 randomly selected countries for each continent. Then arrange the continent names in reverse order. **Hint:** Use the `dplyr` functions `arrange()` and `sample_n()`, they have similar syntax to other dplyr functions.

```{r}
set.seed(234876)
gapminder %>%
  filter(year == 2002) %>%
  group_by(continent) %>%
  sample_n(2) %>%
  summarize(mean_le = mean(lifeExp)) %>%
  arrange(desc(mean_le))
```

```{r}
gapminder %>%
  filter(year == 2002) %>%
  group_by(continent) %>%
  summarize(mean_le = mean(lifeExp),
    min_le = min(lifeExp),        
    max_le = max(lifeExp),
    se_le = sd(lifeExp) / sqrt(n())
    )

tmp <- mutate(gapminder,
              gdp_billions = gdpPercap * pop / 1e9) %>%
  group_by(continent, year) %>%
  summarize(mean_gdp = mean(gdp_billions),
            sd_dgp = sd(gdp_billions))

# Figure
filter(gapminder, continent == 'Americas') %>%
  ggplot(aes(year, lifeExp, color = country)) +
  geom_line() +
  theme(legend.position = 'none') # +
  # facet_wrap(~ country)

gapminder %>%
  mutate(startsWith = substr(country, 1, 1)) %>%
  filter(startsWith %in% c('A','Z')) %>%
  ggplot(aes(year, lifeExp, color = continent)) +
  geom_line() +
  facet_wrap(~ country)
```

### [Data frames with `tidyr`](https://swcarpentry.github.io/r-novice-gapminder/14-tidyr/index.html)

::: callout-tip
## Key Points

-   Use the `tidyr` package to change the layout of data frames.

-   Use `pivot_longer()` to go from wide to longer layout.

-   Use `pivot_wider()` to go from long to wider layout.
:::

#### Challenge 1

Is gapminder a purely long, purely wide, or some intermediate format? **Both**

#### gap_wide vs gap_long vs gapminder

The `gap_wide` data set is formatted such that there is only one row per country.

```{r}
library(readr)
gap_wide <- read_csv('https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_wide.csv')

gap_wide
```

The `gap_long` data set is formatted such that there is one row per data point.

```{r}
gap_long <- gap_wide %>%
  pivot_longer(
    cols = c(starts_with('pop'), 
             starts_with('lifeExp'),
             starts_with('gdpPercap')),
    names_to = "obstype_year", 
    values_to = "obs_values") %>%
    
    separate(obstype_year, into = c('obstype', 'year'), sep = '_')

gap_long
```

Both formats cover the extreme ends of the spectrum of how we could format this in a tidy way. The intermediate format provided by `data(gapminder)` is probably more useful in most situations.

```{r}
gapminder
```

#### Challenge 2

Using `gap_long`, calculate the mean life expectancy, population, and gdpPercap for each continent. **Hint:** use the `group_by()` and `summarize()` functions we learned in the `dplyr` lesson

```{r}
group_by(gap_long, continent, obstype) %>%
  summarize(means = mean(obs_values))
```

#### Challenge 3

Create a `gap_ludicrously_wide` format data by pivoting over countries, year and the 3 metrics? **Hint** this new data frame should only have 5 rows.

```{r}
gap_long %>%
  unite(id_var, country, obstype, year, sep ='_') %>%
  pivot_wider(names_from = id_var, values_from = obs_values)
```

```{r}
gap_normal <- gap_long %>%
  pivot_wider(names_from = obstype, values_from = obs_values) %>%
  select(country, continent, year, lifeExp, pop, gdpPercap) %>%
  arrange(country, continent) %>%
  mutate(year = as.integer(year),
         pop = as.integer(pop))

# This is broken, idk why?
# gap_normal %>%
#  unite(id_var, obstype, year, sep = '_') %>%
#  pivot_wider(names_from = id_var, values_from = obs_values)
```
